---
title: "DNA Methylation Deconvolution Protocol - RefFreeEWAS"
author: Michael Scherer, Petr Nazarov, Reka Toth, Shashwat Sahay, Tony Kamoa, Valentin
  Maurer, Christoph Plass, Thomas Lengauer, Joern Walter, and Pavlo Lutsik
date: "March 05, 2020"
output:
  html_document: default
  pdf_document: default
bibliography: bibliography.bib
---

# Installation and Data Retrieval
Pleas follow the instructions provided in the [protocol](protocol.html) using MeDeCom.

# Data import
Analogously to the general protocol, we use [RnBeads](https://rnbeads.org) for data import.

```{r import, eval=FALSE}
suppressPackageStartupMessages(library(RnBeads))
rnb.options(
  assembly="hg19",
  identifiers.column="submitter_id",
  import=T,
  import.default.data.type="idat.dir",
  import.table.separator="\t",
  import.sex.prediction=T,
  qc=T,
  preprocessing=F,
  exploratory=F,
  inference=F,
  differential=F,
  export.to.bed=F,
  export.to.trackhub=NULL,
  export.to.csv=F
)
sample.anno <- "annotation/sample_annotation.tsv"
idat.folder <- "idat/"
dir.report <- paste0("report",Sys.Date(),"/")
temp.dir <- "/DEEP_fhgfs/projects/mscherer/data/450K/TCGA/LUAD/tmp/"
options(fftempdir=temp.dir)
rnb.set <- rnb.run.analysis(dir.reports = dir.report, sample.sheet = sample.anno, data.dir = idat.folder)
```

# Preprocessing and Filtering
The preprocessing and filtering is also performed in line with the general [protocol](protocol.html) using [*DecompPipeline*](https://github.com/CompEpigen/DecompPipeline).

```{r processing, eval=FALSE}
suppressPackageStartupMessages(library(DecompPipeline))
data.prep <- prepare_data(RNB_SET = rnb.set,
                          analysis.name = "TCGA_RefFree",
                          NORMALIZATION = "bmiq",
                          FILTER_BEADS = T,
                          MIN_N_BEADS = 3,
                          FILTER_INTENSITY = T,
                          MIN_INT_QUANT = 0.001,
                          MAX_INT_QUANT = 0.999,
                          FILTER_NA = T,
                          FILTER_CONTEXT = T,
                          FILTER_SNP = T,
                          FILTER_SOMATIC = T,
                          FILTER_CROSS_REACTIVE = T,
                          execute.lump=T,
                          remove.ICA=T,
                          conf.fact.ICA=c("age_at_diagnosis","race","gender","ethnicity"),
                          ica.setting=c("alpha.fact"=1e-5))
```

# Feature selection
We select the 5,000 most variable CpGs across the samples for downstream analysis.

```{r feature_selection, eval=FALSE}
cg_subset <- prepare_CG_subsets(rnb.set=data.prep$rnb.set.filtered,
                                MARKER_SELECTION = "var",
                                N_MARKERS = 5000)
```

# Deconvolution using RefFreeCellMix

We use the *RefFreeCellMix* function from the [RefFreeEWAS](https://cran.r-project.org/web/packages/RefFreeEWAS/index.html) package for deconvolution analysis. In contrast to *MeDeCom*, *RefFreeCellMix* does not use a regularization for the entries of the matrix to be at the extreme values 0 or 1. Thus, we just specify the number of components to be tested and the preprocessed object, along with the selected features.

```{r deconvolution, eval=FALSE}
md.res <- start.refreeewas.analysis(
  rnb.set=data.prep$rnb.set.filtered,
  cg_groups = cg_subset,
  Ks=2:15,
  factorviz.outputs=TRUE,
  work.dir = "TCGA_RefFree"
)
```

# Downstream analysis

The resulting object is an object of type *MeDeComSet*, since the output from *RefFreeCellMix* has internally been transformed to the *MeDeCom* infrastructure. This enables loading the data set directly into the visualization tool [*FactorViz*](https://github.com/CompEpigen/FactorViz)

```{r visualization, eval=FALSE}
suppressPackageStartupMessages(library(FactorViz))
startFactorViz(file.path(getwd(),"TCGA_RefFree","FactorViz_outputs"))
```